### 2013-04-18 article.tex
Add final reviewer requested revisions
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. In contrast there remains a
steep learning curve in the subsequent computational steps to required to
process raw sequence reads into a database-deposited genome sequence. Here we
describe ``Genomer,'' a tool that simplifies the manual tasks of finishing and
uploading a genome sequence to a database. Genomer can format a genome scaffold
into the common files required for submission to GenBank. Genomer simplifies
editing a genome scaffold by allowing a human-readable YAML format file to be
edited instead of large sequence files. Genomer is written as a command line
tool and is an effort to make the manual, and increasingly common, process of
genome scaffolding more robust and reproducible. Extensive documentation and
video tutorials are available at http://next.gs.

@@

  \item[Simple editting of a draft genome sequence.]{Genomer is built on the
  existing Scaffolder \cite{barton2012} file format for assembling a draft
  genome sequence. This YAML \cite{yaml} format simplifies the process of
  re-organising and trimming contigs in a draft genome without having to edit
  large nucleotide sequences manually.}

  \item[Simple mapping of annotations to the generated draft sequence.]{Genomer
  maps the coordinates of annotations on individual to repsective position in
  the draft genome sequence. This allows the scaffold file to be continuously
  improved even when already annotated. For instance, additional PCR sequences
  can be used to close gaps and improve the genome sequence, without requiring
  reannotation or manually updating all annotation coordinates.}

  \item[Generation of files for submission to GenBank.]{Submitting a genome
  sequence to GenBank database requires several specific files to be prepared.
  These may include a FASTA file of the draft sequence, a table of annotations,
  a FASTA file containing the individual contig sequences, and `a golden path'
  (AGP) \cite{agp-spec} file describing the placement of contigs. Genomer
  automates the generation of all these files.}

  \item[Integration with the command line.]{Genomer is built as a command line
  tool built around plain text files following the Linux philosophy. This
  allows for integration with tools such as GNU Make or git, and also to allow
  reproducible scripting of the process of finishing a genome.}

@@

This software was written for the command-line to use in shells such as bash or
zsh. Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the \verb+make+ command. Additionally we stored each version of the GNU
Make build file with git \cite{git-scm} to track changes to the project and to
allow errors to be reverted from earlier versions. Example build files from our
genome projects can be found on GitHub for a simple plasmid
\cite{plasmid-github} and a more complex genome \cite{genome-github} sequence.

We divided this tool into modular parts each called a `plugin.' Genomer
provides no functionality itself, but rather serves to make the functionality
of plugins available for use at the command line. For example to validate the
correctness of genome annotations, the genomer-plugin-validate plugin should be
installed which then allows the required functionality to be through the
\verb+genomer+ command. Presently, we have developed and released three
plugins: viewing a genome sequence in common bioinformatics file formats,
validating the correctness of genome annotations for common errors, and provide
summaries of a genome sequence. Each of these plugins are available on GitHub
and their use described is in the documentation at http://next.gs.

The Genomer plugin system was built using the existing Ruby package management
system: RubyGems and Bundler. This allows plugins to be automatically installed
without requiring manual downloading and compilation by the user. We anticipate
this will eliminate a common problem where software may require technical
expertise to first compile and install before being used. A plugin system also
allows each plugin to be developed on different release schedules without
requiring Genomer itself to be updated, thereby providing a stable interface.
Each Genomer plugin can be locked to specific versions to prevent changes in
functionality between different plugin versions. This addresses situations
whereby backwards incompatible software changes can break an existing project's
workflow.

@@

\subsection*{First Use Case: Continuous Genome Finishing}

Genome annotation is required to identify the features responsible for
phenotype. Online tools such as the Integrated Microbial Genomes system
\cite{markowitz2006} provide an online service whereby a nucleotide sequence
may be submitted and a subsequent file of predicted genomic features returned.
These annotations are required when submitting a genome sequence to a database.
Once annotated modifying the sequence becomes problematic as this requires
corresponding changes in all downstream annotated coordinates. This therefore
presents a problem whereby either no more updates can be made to the sequence
or the delta difference in annotation coordinates must be propagated to all
downstream locations. The former option is often chosen as being simpler and
therefore the genome sequence must be considered `frozen' once annotated and no
more changes can be made.

During our own \emph{P. fluorescens} genome sequencing projects, we aimed to
continuously improve the quality of the genome sequence by closing gaps using
PCR and Sanger sequencing while at the same time analysing the observed genomic
features. We therefore incorporated functionality into Genomer to allow
annotation coordinates to be automatically updated following such changes to
the sequence. Genomer allows this by organising a genome project as follows:

@@

\item{A genomer project is organised around a set of already assembled and
annotated contigs. For example we assemble our contigs using the A5 pipeline
\cite{tritt2012} and and then subsequently annotate them using the Integrated
Microbial Genomes resource \cite{markowitz2006}. We then use these contigs as a
starting point to build a draft genome sequence in genomer.}

\item{A genome scaffold is then created using the scaffolder file format
\cite{barton2012}. This specifies the order and orientation of contigs in the
scaffold allowing any unresolved gap regions to also be specified. Paired-end
sequencing, comparison to reference genomes, or PCR amplification can be used
to estimate the positions of observed contigs in the true genome sequence.}

\item{The genomer-plugin-view plugin is then added to the Gemfile in the root
of the project directory. The command \verb+bundle update+, run in the project
root, will then download and install this plugin making it available for use in
the Genomer project.}

\item{Different views can then be generated from the scaffold using the
\verb+genomer view+ command. For instance, a GFF3 file of the scaffolded
annotations may be generated. Should the originating contig annotations or the
genome scaffold be changed then rerunning the \verb+genomer view+ command will
correspondingly update the annotation coordinates for the scaffold.}

@@

\subsection*{Second Use Case: Preparing files for submission to GenBank}

The genomer-view-plugin can be used to generate for submission to GenBank:
assembled FASTA, contig FASTA, annotation tables, and scaffold layout in AGP.
This includes FASTA sequences for both the scaffolded and individual contigs, a
GenBank table file of contig annotations, and an AGP file for the order of
contigs in the scaffold. Additional options may also be used to add a prefix to
gene annotation IDs, or begin locus ID numbering at the origin of the sequence,
thereby simplifying the process of creating the required files for submission.

Genomer does not produce the .asn file which must be submitted to GenBank,
instead the generated files are used as the input for the tbl2asn or sequin
tools. We emphasise that the purpose of Genomer is not to reinvent the
functionality already provided by current NCBI resources, but instead to
simplify the process of producing the required inputs. Our example \emph{P.
fluorescens} GNU Make file \cite{genome-github} illustrates how Genomer may be
combined with tbl2asn to produce the required .asn file for submission.

>>>>

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. In contrast there remains a
steep learning curve in the subsequent computational steps required to process
raw sequence reads into a database-deposited genome sequence. Here we describe
``Genomer,'' a tool that simplifies the manual tasks of finishing and uploading
a genome sequence to a database. Genomer can format a genome scaffold into the
common files required for submission to GenBank. Genomer simplifies editing a
genome scaffold by allowing a human-readable YAML format file to be edited
instead of large sequence files. Genomer is written as a command line tool and
is an effort to make the manual, and increasingly common, process of genome
scaffolding more robust and reproducible. Extensive documentation and video
tutorials are available at http://next.gs.

@@

  \item[Simple editing of a draft genome sequence.]{Genomer is built on the
  existing Scaffolder \cite{barton2012} file format for assembling draft genome
  sequences. This YAML \cite{yaml} format simplifies the process of
  re-organising and trimming contigs in draft genome without having to edit
  large nucleotide sequences manually.}

  \item[Mapping of annotations onto the assembled sequence.]{Genomer maps the
  coordinates of contig annotations to respective positions in the draft genome
  sequence. This allows the scaffold file to be continuously improved and
  updated even after annotation. For instance, additional PCR sequences can be
  used to close gaps in the assembly without requiring reannotation or manually
  updating annotation coordinates.}

  \item[Generation of files for submission to GenBank.]{Submitting a genome
  sequence to GenBank database requires generating specific files. These may
  include a FASTA file of the draft sequence, a table of annotations, a FASTA
  file containing the individual contig sequences, and `a golden path' (AGP)
  \cite{agp-spec} file describing the placement of contigs. Genomer automates
  the generation of all these files from the assembled draft genome.}

  \item[A stable interface and streamlined install process]{Genomer provides a
  plugin system built using the existing Ruby package management system:
  RubyGems and Bundler. This allows genomer and its plugins to be automatically
  installed without requiring manual downloading and compilation by the user.
  This eliminates a common problem in bioinformatics where software may require
  technical expertise to first compile and install before being used. Plugins
  can also be locked to specific versions to prevent backwards incompatible
  software changes breaking an existing workflow.}

  \item[Integration with the command line.]{Genomer is built as a command line
  tool built around plain text files. This allows for integration with common
  Linux tools tools such as GNU Make or git. This thereby allow reproducible
  scripting of the genome finishing process.}

@@

As Genomer is written the command-line to use in shells such as bash or zsh.
Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the \verb+make+ command. This is an advantage of command-line tools
versus GUI-based tools, the latter are harder to automate and share steps taken
to produce results. Therefore Genomer may be of particular interest to
bioinformaticians who prefer automated approaches and scripting on the command
line.

Additionally as Genomer uses plain text it is easy to store versions the
project with revision control such as git \cite{git-scm}. This allows tracking
changes to the project and to reverting errors back to earlier versions.
Example build files from our genome projects can be found on GitHub for a
simple plasmid \cite{plasmid-github} and a more complex genome
\cite{genome-github} sequence.

@@

\subsection*{Example use case with Genomer}

Online tools such as the Integrated Microbial Genomes system
\cite{markowitz2006} provide an online service whereby a nucleotide sequence
may be submitted and a subsequent file of predicted genomic features returned.
Once annotated modifying the sequence becomes problematic as this requires
corresponding changes in all downstream annotated coordinates. This therefore
presents a problem whereby either no more updates can be made to the sequence
or the delta difference in annotation coordinates must be propagated to all
downstream locations.

During our own \emph{P. fluorescens} genome sequencing projects, we aimed to
continuously improve the quality of the genome sequence by closing gaps using
PCR and Sanger sequencing while at the same time analysing the observed genomic
features. We therefore incorporated functionality into Genomer to allow
annotation coordinates to be automatically updated following such changes to
the sequence.

@@

  \item{A Genomer project is organised around a set of already assembled and
  annotated contigs. For example we assemble our reads into contigs using the
  A5 pipeline \cite{tritt2012} and and then subsequently annotate them using
  the Integrated Microbial Genomes resource \cite{markowitz2006}. We then use
  these contigs as a starting point to build a draft genome sequence in
  Genomer.}

  \item{A genome scaffold is written in the scaffolder file format
  \cite{barton2012}. This specifies the order and orientation of contigs in the
  scaffold. Any unresolved gap regions can also be specified in this file.
  Paired-end sequencing or comparison to reference genomes are a source of
  contig orientation and order. In our genome projects we determine contig
  order by aligning contigs to a reference genome using nucmer
  \cite{kurtz2004}.}

  \item{The \verb+genomer view+ command can be used to generate files of the
  assembled scaffold. These may include the FASTA file of the assembly or a
  GFF3 file of the annotations in the assembled genome. These files can then be
  used for downstream analysis of the sequence.}

  \item{The genome scaffold can be continuously updated through closing any
  remaining gaps. During our microbe sequencing projects we closed gaps through
  a combination of Sanger PCR and further detailed \emph{in silico} analysis.
  Any changes to the scaffold file automatically propagate to generated files
  when the \verb+genomer view+ command is rerun.}

  \item{When researchers are satisfied with the status of their draft the
  required files for submission to GenBank can be generated: assembled FASTA,
  contig FASTA, annotation tables, and scaffold layout in AGP. Additional
  options are used to add the required prefix to gene annotation IDs and begin
  locus ID numbering at the origin of the sequence. Genomer does not produce
  the .asn file which must be submitted to GenBank, but instead the files for
  the input to tbl2asn or sequin. }

@@

\subsection*{Limitations}

Genomer was written to satisfy our needs for building and finishing a draft
microbial genomes less than 10MBp in size. Draft sequence assembly takes place
in memory and therefore building very large sequences in memory may limit
performance. We point out however that sizes of even 3GBp should fit in the
memory of a modern laptop or desktop computer.

Genomer simplifies the process of generating the annotation table file required
for submission to GenBank where only gene entries are required. Specifically a
user need only provide a GFF3 file of annotations and Genomer can automatically
generate the additional encoded protein features. This however assumes no
intron/exon structure and therefore users wishing to submit annotations with
alternative splicing cannot take advantage of this genomer feature.
